---
title: 'XML Parsing and Database Integration'
author: "Shashank Bettada"
date: "20 April 2023"
output:
  pdf_document: default
  html_document: default
---

Loading the XML library that is required for this assignment.
```{r, include = FALSE, warning = FALSE}

library(XML)
library(RSQLite)
library(stringr)

```

Reading the external XML file into R.
```{r}

path <- getwd()

xpath <- "pubmed22n0001-tf.xml"

xmlFile = paste(path, xpath, sep = "/")

xmlDoc <- xmlParse(file = xmlFile, validate = T)

```

```{r}

fpath = "./"

dbfile = "practicum2DB.sqlite"

dbcon <- dbConnect(RSQLite::SQLite(), paste0(fpath, dbfile))

```

```{r}

dbExecute(dbcon, "DROP TABLE IF EXISTS journals;")

```

```{r}

dbExecute(dbcon, "CREATE TABLE journals (
                  id INTEGER,
                  issn TEXT,
                  title TEXT,
                  primary key (id)
                  );")

```

```{r}

dbExecute(dbcon, "DROP TABLE IF EXISTS author_article;")

```

```{r}

dbExecute(dbcon, "CREATE TABLE author_article (
                  pmid INTEGER,
                  aid INTEGER,
                  primary key (pmid, aid),
                  foreign key (pmid) references articles(pmid),
                  foreign key (aid) references articles(aid)
                  );")

```



```{r}

dbExecute(dbcon, "DROP TABLE IF EXISTS articles;")

```

```{r}

dbExecute(dbcon, "CREATE TABLE articles (
                  pmid INTEGER,
                  journal_id INTEGER,
                  title TEXT,
                  pub_date TEXT,
                  primary key (pmid),
                  foreign key (journal_id) references journals(id)
                  );")

```


```{r}

dbExecute(dbcon, "DROP TABLE IF EXISTS authors;")

```


```{r}

dbExecute(dbcon, "CREATE TABLE authors (
                  aid INTEGER,
                  name TEXT,
                  primary key (aid)
                  );")

```


```{r}

r <- xmlRoot(xmlDoc)

```


```{r}

numArticles <- xmlSize(r)
numArticles

```

Creating a data frame for articles.
```{r}

article_df <- data.frame(pmid = integer(),
                         journal_id = integer(),
                         title = character(),
                         pubDate = character(),
                         stringsAsFactors = F)


```

```{r}

journal_df <- data.frame(jid = integer(),
                         jissn = character(),
                         title = character(),
                         stringsAsFactors = F)


```


```{r}

author_df <- data.frame(aid = integer(),
                        full_name = character(),
                        stringsAsFactors = F)


```

```{r}

author_article_df <- data.frame(pmid = integer(),
                                aid = integer(),
                                stringsAsFactors = F)


```



```{r}

months_dict <- c(
  "Jan" = "01", 
  "Feb" = "02", 
  "Mar" = "03",
  "Apr" = "04", 
  "May" = "05", 
  "Jun" = "06",
  "Jul" = "07", 
  "Aug" = "08", 
  "Sep" = "09",
  "Oct" = "10", 
  "Nov" = "11", 
  "Dec" = "12"
)

seasons_dict <- c(
  "Spring" = "03", 
  "Summer" = "05", 
  "Fall" = "09",
  "Winter" = "12"
)

```

```{r}

# Helper function to extract date
extractDate <- function(journalNode) {
  
  # Trial code begins
  
  year <- NULL
  month <- NULL
  season <- NULL
  day <- NULL
  medline <- NULL
  
  medline <- xpathSApply(journalNode, "./Journal/JournalIssue/PubDate/MedlineDate", xmlValue)

  if (length(medline) == 0) {
    
    year <- xpathSApply(journalNode, "./Journal/JournalIssue/PubDate/Year", xmlValue)
  
    month <- xpathSApply(journalNode, "./Journal/JournalIssue/PubDate/Month", xmlValue)
    if (length(month) == 0) {
      month <- NULL
      season <- xpathSApply(journalNode, "./Journal/JournalIssue/PubDate/Season", xmlValue)
      if (length(season) == 0) {
        season <- NULL
      }
      
    }
    
    day <- xpathSApply(journalNode, "./Journal/JournalIssue/PubDate/Day", xmlValue)
    if (length(day) == 0) {
      day <- NULL
    } 
  
    medline <- NULL
  } 
  
  
  # Trial code ends
  
  year_ <- "0000"
  month_ <- "Xxx"
  day_ <- "00"
  
  
  if (!is.null(medline)) {
    
    # Check year
    year_ <- substr(medline, 1, 4)
    
    # Check month
    mon <- substr(medline, 10, 12)
    month_ <- months_dict[mon]
    
    # Check day
    day_ <- "01"

  } else {
    
    # Check year
    year_ <- year
    
    # Check month
    if (!is.null(month)) {
      month_ <- months_dict[month]
    } else if (!is.null(season)) {
      month_ <- seasons_dict[season]
    } else {
      month_ <- "01"
    }
    
    # Check day
    if (!is.null(day)) {
        day_ <- day
      } else {
        day_ <- "01"
      }

  }
  
  # return (paste0(month_, "-", day_, "-", year_))
  return (paste0(year_, "-", month_, "-", day_))

}

```



```{r}

fetchNameAndKey <- function(authorNode) {
  
  lname <- xpathSApply(authorNode, "./LastName", xmlValue)
  
  if (length(lname) == 0) {
    collective <- xpathSApply(authorNode, "./CollectiveName", xmlValue)
    key <- str_replace_all(collective, " ", "-")
    nameKey <- key
    
    name <- collective
    
  } else {
    
    fname <- xpathSApply(authorNode, "./ForeName", xmlValue)
    # initial <- xpathSApply(authorNode, "./Initials", xmlValue)
    
    if (length(fname) != 0) {
      nameKey <- paste0(lname, fname)
      name <- paste0(lname, ", ", fname)
    } else {
      nameKey <- lname
      name <- lname
    } 

  }
  
  df <- c(nameKey, name)
  
  return(df)
}

```

```{r}

journal_count <- 1
journal_unknown_count <- 1
author_count <- 1
author_article_count <- 1

```


```{r}

# Create a new df in order to keep track of repeated journals
j_issn <- c("")

j_id <- c(1)

j_title <- c("")

journalCheck_df <- data.frame(j_issn, j_id, j_title)

journalCheck_df

```


```{r}

# Create a new df in order to keep track of repeated journals
a_key <- c("")

a_id <- c(1)

authorCheck_df <- data.frame(a_key, a_id)

authorCheck_df

```


```{r}

# Parse article

parseArticle <- function(articleNode) {
  

  # Article elements
  attr <- xmlAttrs(articleNode)
  pmid <- as.integer(attr[[1]])
  
  journal_title <- xpathSApply(articleNode[[1]], "./Journal/Title", xmlValue)
  
  journal_issn <- xpathSApply(articleNode[[1]], "./Journal/ISSN", xmlValue)
  if (length(journal_issn) == 0) {
    
    if (any(journalCheck_df$j_title == journal_title) == FALSE) { 
      journal_issn <- paste0("unknown-", journal_unknown_count)
      journal_unknown_count <<- journal_unknown_count + 1
    } else {
      
      journal_issn <- journalCheck_df$j_issn[which(journalCheck_df$j_title == journal_title)]
    }
  }

  # Check if the journal is already present in the df
  if (any(journalCheck_df$j_issn == journal_issn) == FALSE) {
    
    #Add journal (Insert new journal to journal_df)
    journalCheck_df[journal_count, ] <<- c(journal_issn, journal_count, journal_title)
    
    journal_df[journal_count, ] <<- c(journal_count, journal_issn, journal_title)

    journal_count <<- journal_count + 1
    
  } 
  
  authorList <- xpathSApply(articleNode[[1]], "./AuthorList/Author")
  
  for (author in authorList){
    
    auth_res <- fetchNameAndKey(author) 
    
      # Check if author is already present in the df
    if (any(authorCheck_df$a_key == auth_res[1]) == FALSE) {
      
      #Add author (Insert new journal to journal_df)
      authorCheck_df[author_count, ] <<- c(auth_res[1], author_count)
      
      author_df[author_count, ] <<- c(author_count, auth_res[2])
      
      author_count <<- author_count + 1
      
    } 
    
    ## Add to author_article_df
    author_id <- which(authorCheck_df$a_key == auth_res[1])
    author_article_res <- c(pmid, author_id)
    author_article_df[author_article_count, ] <<- as.character(author_article_res)
    author_article_count <<- author_article_count + 1
    
  }
  
  journal_id <- which(journalCheck_df$j_issn == journal_issn)

  title <- xpathSApply(articleNode, "./PubDetails/ArticleTitle", xmlValue)
  
  # Extract article date 
  article_date <- extractDate(articleNode[[1]])
  
  df <- c(pmid, journal_id, title, article_date)
  
  return(df)

}


```


```{r}

# Insert data to articles df

# Took 5 minutes to parse.

for (i in 1:numArticles) {
  
  x <- xmlRoot(xmlDoc)
  
  article_df[i, ] <- parseArticle(x[[i]])
  
}


```

```{r}

head(article_df)

```

```{r}

head(journal_df)

```

```{r}

head(author_df)


```

```{r}

head(author_article_df)

```


```{r}

save(article_df, file = "article.Rdata")
save(journal_df, file = "journal.Rdata")
save(author_df, file = "author.Rdata")
save(author_article_df, file = "author_article.Rdata")
save(journalCheck_df, file = "journalCheck.Rdata")
save(authorCheck_df, file = "authorCheck.Rdata")

```

```{r, message = FALSE, warning = FALSE}

dbWriteTable(dbcon, "articles", article_df, row.names = F, overwrite = T)
dbWriteTable(dbcon, "journals", journal_df, row.names = F, overwrite = T)
dbWriteTable(dbcon, "authors", author_df, row.names = F, overwrite = T)
dbWriteTable(dbcon, "author_article", author_article_df, row.names = F, overwrite = T)

```


```{sql connection = dbcon}

SELECT Count(*) FROM articles;

```

```{sql connection = dbcon}

SELECT * FROM articles
LIMIT 5;

```


```{sql connection = dbcon}

SELECT articles.pubDate FROM journals
JOIN articles ON (articles.journal_id = journals.jid)
WHERE journals.jissn = "0006-2944";

```

```{sql connection = dbcon}

SELECT * FROM journals
LIMIT 20;

```



```{sql connection = dbcon}

SELECT * FROM authors
LIMIT 5;

```


```{sql connection = dbcon}

SELECT * FROM author_article
LIMIT 5;

```

```{sql connection = dbcon}

SELECT j.jid, j.jissn, j.title, strftime('%Y', date(a.pubDate)) as Year, strftime('%m', date(a.pubDate)) as Month, count(distinct(a.pmid)) as Articles, count(distinct(au.aid)) as Authors
  FROM journals j
  LEFT JOIN articles a
  ON j.jid = a.journal_id
  LEFT JOIN author_article aau
  ON a.pmid = aau.pmid
  LEFT JOIN authors au
  ON aau.aid = au.aid
  GROUP BY j.jid, Year, Month
  ORDER BY j.jid
  LIMIT 10;

```

## Part 2

```{r}

library(tibble)

```

```{r}

facts_query <- "SELECT j.jid, j.jissn, j.title, strftime('%Y', date(a.pubDate)) as Year, 
                strftime('%m', date(a.pubDate)) as Month, count(distinct(a.pmid)) as Articles, 
                count(distinct(au.aid)) as Authors
                  FROM journals j
                  LEFT JOIN articles a
                  ON j.jid = a.journal_id
                  LEFT JOIN author_article aau
                  ON a.pmid = aau.pmid
                  LEFT JOIN authors au
                  ON aau.aid = au.aid
                  GROUP BY j.jid, Year, Month
                  ORDER BY j.jid;"

```

```{r}

facts_res_df <- dbGetQuery(dbcon, facts_query)

```

```{r}

head(facts_res_df) 

```

```{r}

quarters_dict <- c(
  "01" = "01", 
  "02" = "01", 
  "03" = "01",
  "04" = "02", 
  "05" = "02", 
  "06" = "02",
  "07" = "03", 
  "08" = "03", 
  "09" = "03",
  "10" = "04", 
  "11" = "04", 
  "12" = "04"
)

```

```{r}

facts_res_df <- add_column(facts_res_df, quarters_dict[facts_res_df$Month], .after = 5)
head(facts_res_df)

```

```{r}


colnames(facts_res_df)[6] = "Quarter"
head(facts_res_df)

```


```{r}

n_facts <- nrow(facts_res_df)

```


```{r}

factId <- matrix("1", ncol = 1, nrow = n_facts)
factId_df <- data.frame(factId)
nrow(factId_df)

```

```{r}

facts_df <- cbind(factId_df, facts_res_df)
head(facts_df)

```


```{r}

facts_df[,1] <- seq(1, n_facts)


```

```{r}

facts_df$jid <- as.integer(facts_df$jid)
facts_df$Year <- as.integer(facts_df$Year)
facts_df$Month <- as.integer(facts_df$Month)
facts_df$Quarter <- as.integer(facts_df$Quarter)

```

```{r}

head(facts_df)

```


```{r}

library(RMySQL) 

```

Configuring the db connection specifications:
```{r db_configuration, message = FALSE, warning = FALSE}

db_user <- 'root'
db_password <- 'ENTER YOUR PASSWORD'
db_name <- 'factsDB'
db_host <- 'localhost'
db_port <- 3306

```

Establishing connection with the MySQL database (localhost):
```{r db_connection, message = FALSE, warning = FALSE}

dbmySql <- dbConnect(MySQL(), user = db_user, password = db_password, 
                  dbname = db_name, host = db_host, port = db_port)

```

```{sql drop_airports, connection = dbmySql}

DROP TABLE IF EXISTS journal_facts;

```

```{sql create_airports, connection = dbmySql}

CREATE TABLE journal_facts (
  factId INT,
  journalId INT,
  issn VARCHAR(10),
  title VARCHAR(150),
  year INT,
  month INT,
  quarter INT,
  articles INT,
  authors INT,
  PRIMARY KEY (factId)
);

```

Loading the data into the incidents table using dbWriteTable() function:

```{r load_incident, message = FALSE, warning = FALSE}

dbWriteTable(dbmySql, 'journal_facts', facts_df, row.names = F, overwrite = T)

```


```{sql create_airports, connection = dbmySql}

ALTER TABLE journal_facts MODIFY factId INTEGER;

```

```{sql create_airports, connection = dbmySql}

ALTER TABLE journal_facts MODIFY jid INTEGER;

```

```{sql create_airports, connection = dbmySql}

ALTER TABLE journal_facts MODIFY Year INTEGER;

```

```{sql create_airports, connection = dbmySql}

ALTER TABLE journal_facts MODIFY Month INTEGER;

```

```{sql create_airports, connection = dbmySql}

ALTER TABLE journal_facts MODIFY Quarter INTEGER;

```

```{sql create_airports, connection = dbmySql}

ALTER TABLE journal_facts MODIFY Articles INTEGER;

```

```{sql create_airports, connection = dbmySql}

ALTER TABLE journal_facts MODIFY Authors INTEGER;

```

```{sql create_airports, connection = dbmySql}

SELECT * FROM journal_facts
WHERE Year = 1977
LIMIT 5;

```

Disconnect SQLite database:
```{r disconnect_db, message = FALSE, warning = FALSE}

dbDisconnect(dbcon)

```

Disconnect MySQL database:
```{r disconnect_db, message = FALSE, warning = FALSE}

dbDisconnect(dbmySql)

```









